Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-04-19 doc/invt.tex
further small edits to invt.tex

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme (WL or metadynamics),
  and the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of sampling steps with a proper shift,
  as given by the $1/t$ prescription.
  %
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p^*(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complex, multi-modal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, *lee1993,
%wang2001, *wang2001pre,
%huber1994,
%*laio2002, *laio2008, *barducci2011, *sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012},
more commonly used for MC and MD simulations, respectively\cite{junghans2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of $z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
Although the assignment of the window functions to the two methods
is not absolute\cite{micheletti2004, kim2006, *kim2007, junghans2014},
below we will use the terms WL and metadynamics
to represent FDS simulations using the
$\delta$- and Gaussian window functions, respectively.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help fix our notations,
and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses metadynamics
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix},
and present a general procedure of computing the optimal schedule
in Sec. \ref{sec:procedure}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

Note that the error weighting distribution, $\pmb\rho$,
defined above
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$.
%
In practice, $\pmb\rho$ is usually flat,
i.e. $\rho_i \equiv 1/n$,
even when $\mathbf p$ is not
(e.g. in well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$).
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

Next, we split $h_i(t)$ into a deterministic averaged part
and a stochastic part, $\zeta_i(t)$,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an ``ensemble average'',
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
and we will assume it is white, i.e.
%The noise is not necessarily white,
%and the autocorrelation function
%depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  %\sigma_{ij}(t - t')
  G^{(0)}_{ij} \, \delta(t - t')
  ,
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
for some value of $G^{(0)}_{ij}$,
and $\delta(t)$ is the Dirac's delta function.
%where $\sigma_{ij}(t)$ is an even function
%that vanishes at large $t$.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of noise.
%
For non-white noise,
the equivalent value of $G^{(0)}_i$ can be approximately obtained
from the integral,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$,
or in the discrete time case,
$\sum_{-\infty}^\infty \sigma_{ij}(t)$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
%
where
%
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = G_i \, \delta(t - t'),
$$
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

@@

Since the above optimization is performed
under a fixed $q(T)$,
it can be improved
by choosing a proper value of $q(T)$.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
following the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
%In Sec. \ref{sec:finlencorr},
%we show that % $t_0$ can be adjusted
%one can further improve the schedule by adjusting $t_0$
%to accommodate the initial error.
%
%and
%$$
%\left\langle x^2(T) \right\rangle
%=
%\frac{ \left\langle x^2(0) \right\rangle \, t_0^2 }
%     { (t_0 + T)^2 }
%+
%\frac{ \Gamma \, T}
%     { (t_0 + T)^2 }
%.
%$$
%Using this result for each $x_i$,
%we get the final total error, $E(T)$,
By using the optimal schedule in Eq. \eqref{eq:ET_average},
we find the final total error to be
\begin{equation}
  E(T)
  =
  \frac{ E(0) \, t_0^2 }
       { (t_0 + T)^2 }
  +
  \frac{ \Gamma \, T}
       { (t_0 + T)^2 }
  ,
  \label{eq:error_singlebin}
\end{equation}
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}

@@

We now consider a multiple-bin updating scheme,
given in Eq. \eqref{eq:mbin_update}.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential at bin $j$,
but also those at a few neighboring bins, $i$'s.
%
The relative updating magnitudes,
$w_{ij}$'s,
form a matrix, $\mathbf w$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
%
\begin{equation}
  \dot V_i(t)
  =
  \alpha(t) \sum_{j=1}^n \, w_{ij} \frac{ h_j(t) } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

The optimization procedure is similar to the single-bin case.
%
The key simplification comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$.
%However, we will assume some restrictions and approximations
%to simplify the discussion.

@@

The matrix, $\mathbf w$, is subject to three conditions.
%
First, to sample
the desired distribution,
$\mathbf p = (p_1, \dots, p_n)$,
${\dot V}_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$
if $h_j(t)$ were the same as $p_j$.
%
This allows the bias potential to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}.
%
The quoted results were unaffected by the allowance
of negative elements into the matrix, $\mathbf w$.

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e. $\pmb \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.
%
In the way,
the error weighting distribution, $\pmb \rho$,
restrains the form of the updating matrix, $\mathbf w$.
%
The condition, albeit somewhat artificial,
is readily satisfied in practice:
most commonly, $\pmb \rho$ is a uniform distribution,
and this condition requires the updating matrix, $\pmb w$, to be symmetric.

With Eq. \eqref{eq:w_detailedbalance},
the scaled matrix, $\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
is symmetric.
Thus, we can diagonalize $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  .
\label{eq:eig_orthonormal_rows}
\end{align}

In particular, there is always an eigenvector
corresponding to the distribution, $\pmb \rho$.
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{1i} = \rho_i,
\label{eq:eigenmode1}
\end{equation}
%
which satisfies Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 1$.
%
The corresponding eigenvalue is $\lambda_1 = 1$
by Eq. \eqref{eq:w_balance}.
%
Further, for $k \ge 2$, we have
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \sum_{ i = 1 }^n
    \frac{ \phi_{ki} \, \phi_{1i} }
         { \rho_i }
  =
  0
  .
\label{eq:ortho1}
\end{equation}

@@

\subsubsection{Linearization and decoupling to eigenmodes}

@@

Since the eigenvalues are related to the updating kernel
by a cosine transform,
Eq. \eqref{eq:wband_eigenvalue},
the relation can be readily inverted.
%
In the periodic case,
we have
%the inversion formula is
%
\begin{equation}
  \mu_l
  =
  \frac 1 n
  \sum_{ k = 1 }^n
  \lambda_k
  \cos \frac{ (k - 1) \, l \, 2 \, \pi }
            {            n             }
  .
\label{eq:mu_from_lambda_pbc}
\end{equation}

>>>>

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme
  (such as the single-bin scheme commonly used in WL
  or the Gaussian one popular in metadynamics),
  and the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of sampling steps with a proper shift,
  as given by the $1/t$ prescription.
  %
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry
that can provide insight into an array of phenomena not easily studied
with traditional experiments.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable, $z$, by sampling via either Monte Carlo\cite{
  frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
The dimensionless free energy, or potential of mean force (PMF),
along the collective variable, $z$,
is $-\ln p^*(z)$.
%
However, accurately estimating the PMF is often difficult
when the collective variable is energetically restricted to local regions
of the complex, multimodal free energy surface.
%
Thus,
to overcome these energetic barriers and
to capture the global shape of the free energy surface
an effective strategy is to introduce a bias potential that
cancels the PMF.
%
Ideally, then, sampling yields a flat distribution
along $z$\cite{mezei1987, berg1992, *lee1993,
wang2001, *wang2001pre,
huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}.
%
The flat-distribution sampling can be used to accelerate
sampling of slow transitions that otherwise would not readily
be observed with traditional simulations.
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, *lee1993,
%wang2001, *wang2001pre,
%huber1994,
%*laio2002, *laio2008, *barducci2011, *sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

Many efficient flat-distribution-sampling (FDS) techniques
based on adaptive construction of the bias potential
have been introduced,
including the Wang-Landau (WL) algorithm\cite{
  wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012},
popular for MC and MD simulations, respectively\cite{junghans2014}.
%
%flat-distribution-sampling (FDS) techniques
%that build the bias potential adaptively or on the fly.
%
These techniques regularly update the bias potential
to discourage future visits to previously sampled configurations
by incrementally elevating the bias potential,
%and they are closely-related\cite{junghans2014, micheletti2004}.
%
%The bias potential is associated with discrete values of $z$.
%
A key difference lies
in the updating window function,
which specifies
a neighborhood around the current value of
$z$ on the free energy surface
where the updating should occur
as well as the relative updating magnitude.
%
Usually, the window function
takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in WL,
or that of a Gaussian
in metadynamics.
%
Although the two window functions
are not always associated with
the two methods\cite{micheletti2004, kim2006, *kim2007, junghans2014},
we will loosely use the terms WL and metadynamics
to represent FDS simulations using the
$\delta$- and Gaussian window functions, respectively.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence of the estimated PMF
one  has to gradually decrease
the updating magnitude of the bias.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help define our notations,
and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses metadynamics
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix},
and present a general procedure of computing the optimal schedule
in Sec. \ref{sec:procedure}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

Although the above error weighting distribution, $\pmb\rho$,
is usually flat,
i.e. $\rho_i \equiv 1/n$,
in practice,
a non-flat distribution
is sometimes more convenient in analyzing
some multiple-bin updating schemes later on.
%
We also note that $\pmb\rho$
is independent of
the target distribution,
$\mathbf p = (p_1, \dots, p_n)$
(e.g. in well-tempered metadynamics,
$\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
  barducci2008, dama2014}:
$\ln p_i = \ln {p^*_i}/(1+\Delta) + \mathrm{const.}$ for some positive $\Delta$,
whereas $\pmb\rho$ is flat).
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

Next, we split $h_i(t)$ into a deterministic averaged part
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an ``ensemble average'',
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

@@

%
For the histogram fluctuation part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
and we will approximate it as white noise, i.e.
%The noise is not necessarily white,
%and the autocorrelation function
%depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  %\sigma_{ij}(t - t')
  G^{(0)}_{ij} \, \delta(t - t')
  ,
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
for some value of $G^{(0)}_{ij}$,
and $\delta(t)$ is the Dirac's delta function.
%where $\sigma_{ij}(t)$ is an even function
%that vanishes at large $t$.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of noise.
%
For non-white noise,
the equivalent value of $G^{(0)}_i$ can be approximately obtained
from the integral,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$,
or in the discrete time case,
$\sum_{-\infty}^\infty \sigma_{ij}(t)$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
%
where the modified histogram fluctuation
%
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ j = 1 }^n
    \rho_j
    \frac{ \zeta_j(t) }
         {     p_j    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also governed by white noise as is $\zeta_i(t)$,
i.e. we may write
$$
  \langle \xi_i(t) \, \xi_i(t') \rangle = G_i \, \delta(t - t'),
$$
for some $G_i$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ h_j } { p_j }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} \; and \; \eqref{eq:xi_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ j = 1 }^n
      \rho_j \frac{ \langle h_j \rangle } { p_j }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_ave} \;and\; \eqref{eq:rhox_sum}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

@@

Since the above optimization is performed
under a fixed $q(T)$,
it can be improved
by choosing a proper value of $q(T)$.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using the optimal schedule in Eq. \eqref{eq:ET_average},
we find the final total error to be
\begin{equation}
  E(T)
  =
  \frac{ E(0) \, t_0^2 + \Gamma \, T}
       { (t_0 + T)^2 }
  ,
  \notag
  %\label{eq:error_singlebin}
\end{equation}
%
which is minimized if
%
\begin{equation}
t_0 = \frac{ \Gamma }
           { E(0) }
           ,
  \label{eq:t0_singlebin}
\end{equation}
%
and at this point
%
\begin{equation}
  E(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}

@@

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
%
\begin{equation}
  \dot V_i(t)
  =
  \alpha(t) \sum_{j=1}^n \, w_{ij} \frac{ h_j(t) } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

The optimization procedure is similar to the single-bin case.
%
The key simplification comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.
%
%However, we will assume some restrictions and approximations
%to simplify the discussion.

@@

The matrix, $\mathbf w$, is subject to three conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\mathbf p = (p_1, \dots, p_n)$,
${\dot V}_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$
if $h_j(t)$ were the same as $p_j$.
%
This allows the bias potential to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}
(we will only use results that are unaffected by the allowance
of negative elements into $\mathbf w$).

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the error weighting distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vbar_def}.
%
This artificial condition
restrains the form of the updating matrix, $\mathbf w$,
and it requires, for the commonly-used uniform distribution, $\pmb\rho$,
a symmetric updating matrix, $\pmb w$.
%
More generally, only
the scaled matrix, $\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
needs to be symmetric,
and the symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  .
  \label{eq:eig_orthonormal_rows}
\end{align}

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{1i} = \rho_i,
\label{eq:eigenmode1}
\end{equation}
%
with $\lambda_1 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode1}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 1$,
while the orthogonality for $k \ge 2$ demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \sum_{ i = 1 }^n
    \frac{ \phi_{ki} \, \phi_{1i} }
         { \rho_i }
  =
  0
  .
\label{eq:ortho1}
\end{equation}

@@

\subsubsection{Eigenmode decomposition}

@@
